---
title: "data_challenge"
author: "LAT"
date: "2024-02-15"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Frequency tables for research participant diversity

```{r}
# import packages
library(tidyr)
library(dplyr, warn.conflicts=FALSE)
library(janitor, warn.conflicts=FALSE)
```

```{r}
# read in CSVs
az_demo <- read.csv("az_demo.csv")
az_test <- read.csv("az_test.csv")
```

```{r}
# investigate data
colnames(az_demo) 
str(az_demo) 
summary(az_demo) 
```

## 1. Table of names and ages

a.  Separate names using tidyr's separate function

```{r}
az_demo <- separate(az_demo, names, into = c("last_name", "first_name"), sep = ",", remove = FALSE) 
```

b.  Calculate age from dob

```{r}
az_demo$dob <- as.Date(az_demo$dob) 
today <- as.Date(Sys.Date()) 
az_demo$age_yrs <- round(as.numeric(today - az_demo$dob) / 365.25, 2) 
```

c.  Print table

```{r}
table <- head(az_demo[, c("id", "names", "first_name", "last_name", "dob", "age_yrs")], n = 5) 
print(table)
```

## 2. Create a frequency/cross tab table of total participants by diagnosis and URG Status

a.  Determine URG status

```{r}
az_demo$URG <- "" 
az_demo$URG <- ifelse(az_demo$race == "White" & az_demo$ethnicity == "Not Hispanic", "Non-URG", ifelse(az_demo$race == "Other" & az_demo$ethnicity == "Not Hispanic", "Unknown", "URG"))
```

b.  Build table with diagnosis as rows and URG status as columns using dplyr package

```{r}
freq_table <- az_demo %>%
  count(diagnosis, URG) %>%
  spread(URG, n, fill = 0) %>%
  mutate(Total = URG + `Non-URG` + Unknown) %>%
  relocate(URG, .before = Unknown) 
```

c.  Add column and row totals using janitor package

```{r}
freq_table <- adorn_totals(freq_table)
```

d.  Include 4 columns for percent of total for all columns + clean up table

```{r}
# make new cols hold percents
freq_table <- freq_table %>%
  mutate(`Non-URG_percent` = (`Non-URG` / tail(`Non-URG`, 1)) * 100) %>%
  mutate(URG_percent = (URG / tail(URG, 1)) * 100) %>%
  mutate(Unknown_percent = (Unknown / tail(Unknown, 1)) * 100) %>%
  mutate(Total_percent = (Total / tail(Total, 1)) * 100)

# reorder columns
freq_table <- freq_table %>%
  select(diagnosis, `Non-URG`, `Non-URG_percent`, URG, URG_percent, Unknown, Unknown_percent, Total, Total_percent)

# round percent values
freq_table$URG_percent <- round(freq_table$URG_percent, 2)
freq_table$`Non-URG_percent` <- round(freq_table$`Non-URG_percent`, 2)
freq_table$Unknown_percent <- round(freq_table$Unknown_percent, 2)
freq_table$Total_percent <- round(freq_table$Total_percent, 2)

# add percent signs
freq_table$URG_percent <- paste(freq_table$URG_percent, "%", sep = "")
freq_table$`Non-URG_percent` <- paste(freq_table$`Non-URG_percent`, "%", sep = "")
freq_table$Unknown_percent <- paste(freq_table$Unknown_percent, "%", sep = "")
freq_table$Total_percent <- paste(freq_table$Total_percent, "%", sep = "")

# reorder rows + reset index
new_order <- c(1, 4, 2, 3, 6, 5, 7)
freq_table <- freq_table[new_order, ]
row.names(freq_table) <- NULL

# rename columns
freq_table <- freq_table %>%
  rename("URG N"= URG, "Non-URG N" = `Non-URG`, "Unknown N" = Unknown, "Total N" = Total, "Diagnosis" = diagnosis, "URG % of Total" = URG_percent, "Non-URG % of Total" = `Non-URG_percent`, "Unknown % of Total" = Unknown_percent, "Total % of Total" = Total_percent)

freq_table
```

## 3. Score ABC test and calculate age at visit

a.  Score ABC test

```{r}
az_test$sum_score <- rowSums(az_test[, c("abc_01", "abc_02", "abc_03", "abc_04", "abc_05", "abc_06", "abc_07", "abc_08", "abc_09", "abc_10")])
az_test$mean_score <- az_test$sum_score / 10
```

b.  Calculate age at visit

```{r}
az_test$visit_date <- as.Date(az_test$visit_date)
```

```{r}
# merge datasets
az_merged <- merge(az_demo, az_test, by.x = "id")
az_merged$visit_age_yrs <- round(as.numeric(az_merged$visit_date - az_merged$dob) / 365.25, 2)
az_merged <- relocate(az_merged, visit_age_yrs, .before = visit_num)
```

## 4. Latest visit and current age

a.  Gather latest visits with groupby

```{r}
latest_visits <- az_merged %>%
  group_by(id) %>%
  filter(visit_num == max(visit_num)) %>%
  summarise(visit_num)
```

b.  Add participant's current age

```{r}
latest_visits <- az_merged %>%
  group_by(id) %>%
  filter(visit_num == max(visit_num)) %>%
  summarise(visit_num, age_yrs)
```

c.  Print table of first 5 rows for specified columns

```{r}
latest_visits <- az_merged %>%
  group_by(id) %>%
  filter(visit_num == max(visit_num)) %>%
  ungroup()
head(latest_visits[c("id", "visit_num", "visit_date", "visit_age_yrs", "age_yrs", "sum_score")], 5)
```
